Preparing for Potential Closure of European Airspaces due to Re-entering Space Objects

### 2025-05-04 paper.qmd
reference for velocity of debris

# Abstract
<!-- teaser --> The recent decades observed an increasing number of space activities and associated launches. 
The risk of collision between uncontrolled re-entering space debris and aircraft
is  increasing due to both ever-growing number of airborne flights and 
of orbiting bodies (and relevant re-entries). 
<!-- why important --> National and international authorities need to balance safety and economic
impact of adopted measures should a re-entry event be announced, usually on short
notice. <!-- no hard line breaks in abstracts -->
<!-- approach / experiments --> We employ hourly and daily spatial flight densities calculated from
trajectories as recorded by EUROCONTROL's Network Manager and
binned with the Uber H3 geospatial indexing system.  
Impact risk probabilities are then calculated based on previous theoretical
work. 
<!-- results / conclusions --> This work enhances the model by employing using actual flown flight trajectories as a basis for time-varying flight densities rather than relying on proxies such as the population density or ADS-B/Mode S message counts.
These densities combined with a buffered area along the eventual path of
re-entry, can provide mid-air impact probabilities to authorities as an aid
for deciding on safety measures such as airspace closures and/or re-routings.

@@

<!-- teaser opening --><!-- might need a rewrite - first jotting down -->
The quest for space is a driver for technological development.
Key milestones include the first human traval to space and surrounding the globe, the first and now many succeeding successful landings on the moon, and regular operations of space stations in the orbit.
These developments helped to identify options for the wider use of space, benefits, and its commercialisation.
This is showcased by an increasing number of activities aimed to generate value through the exploration of space ranging from scientific understanding and experimentation, sustainable prooduction of goods in a non-gravity environment, services, and recently leisure.
The increased number of launches, however, also increased the number of artefacts <!-- falling down and --> re-entering the atmosphere.

<!-- zooming in on problem -->
With forecasts predicting a steadily growing number of flights, see @fig-forecast, and the modern
race to commercialise space, there is a need to re-evaluate the risks of
collision and casualties due to uncontrolled debris re-entries.
The increasing risk landscape poses challenges in terms of oversight and risk management.
There is a growing demand by overseeing authorities for new methodologies and tools to assess which 
safety measures are the most adequate to minimise potential casualties and economic
impact.

The increase in space operations does not exist in a vacuum. 
The continual growth of air traffic - despite temporal bumps - is evidenced by the surging number of flights in Europe [@forecastUpdate2025] and across the world. 
On-going development/deployment activities of mega-constellations of satellites in Low Earth Orbit \acr{LEO}, 
such as already active Starlink and OneWeb or planned ones (e.g. Kuiper, Starshield, Xingwang) will add thousand of orbiting objects with a short operational life. 
Some of these will inevitably re-enter the Earth's atmosphere in an uncontrolled manner, c.f. @tbl-megaconstellations [@megaconstellations].

@@

Aviation safety risk management is a well-established domain in air transportation. 
Despite the increasing need for situational awareness of space objects and their re-entry, the classical risk avoidance paradigm revolves around restricting the airspace.
While the probability of a strike is low, it can result in catastrophic consequences. 
The growing number of space operations requires NEW - ADVANCED TOOLS. PREVIOUS WORK DID A,B, and MAYBE C - WE DO SOMETHING ON TOP.

@@

<!-- ESTABLISH SOME CONTEXT ---------------------------------------------- -->
The modelling and tracking of uncontrolled re-entries poses challenges.
Actual and historic data is sparse, and the size, shape, and attitude or rotation of re-entering objects increase the level of uncertainty for modelling.
Common approaches employ a mix of handingling the uncertainty of object or debris movement and exposed area, including temporal uncertainty windows.

      <!-- is there more to say and provide context? -->

<!-- APPROACH / METHOD --------------------------------------------------- -->
<!-- it would be cool, if we can add a diagram here that shows the path from data, major processing building blocks to the results -->
<!-- as a minimum, we should say something about the WHAT -->
In general, the method for estimating the risk to commercial aircraft from reentering debris objects revolves around estimating the exposure of a single flight and the population of flights to the re-entering object.
In that respect, a robust determination of flight density is a key aspect.

We follow the same mathematical approach of [@patera2008; @wrightAirspaceClosuresDue2025].
However, this paper augments the approximation of risk exposure. 
Instead of approximating flight density with population density [@patera2008] or
aircraft transponder data [@wrightAirspaceClosuresDue2025], we use actual flown
flight trajectories.

This work focusses on the European region.
The hourly flight density estimation is based on on flown \acr{IFR} trajectories as
recorded by EUROCONTROL's \acr{NM}[^1].
To decrease spatial complexity, the data is binned using Uber's H3 geospatial indexing system [@uber-h3]. 
The reduction of the spatial complexity enables fast processing.
For this work, we employ a H3-cell-resolution of 3. 
This corresponds to hexagons of an average area $\qty{12393}{km^2}$ and an average edge length of
$\qty{68.979}{km}$ [@h3resolutions] as depicted in @fig-cells.

<!-- CAN WE MOTIVATE WHY resolution 3??? -- e.g. is this an area used in other works, or does it do something? -->

@@

To build an upper threshold for flight exposure, this study uses the busiest week during the \acr{IATA} summer ISO week in 2023 and 2024.   <!-- does ISO do here anything? --> 
The flight densities are characterised on an hourly basis for all days in these weeks.

<!-- next sentence feels like an orphan ... do we need this here? ------>
This paper expands the works of [@wrightAirspaceClosuresDue2025; @patera2008] and provides a reproducible example for States, Agencies or relevant authorities to build upon.

@@

<!-- next sentence feels like an orphan ... do we need this here? ------>
Next sections will develop the mathematical background for the evaluation of collision and casualties risk of which traffic density is a key component.

@fig-density-summer2024-res-3 shows hourly flight density for the 5th of July 2024.
The chart shows the expected outcomes in terms of major traffic flows, and areas of high or low traffic.
For example, cells with a density of zero are clearly visible for areas of war or no-fly zones
(Ukraine, Libia or Syria). The major traffic flows (North Atlantic, Canaries
and East-bound over Türkiye) are clearly highlighted showing densities of more than 10.

@@

* constant debris fall speed $s_{d_k} = \qty{145}{mph} = S$
* constant aircraft speed, $s_{a} = C_t$, equal to its cruise speed, 
* aircraft area as viewed from the top, $a_T = W_t \times L_t = A_t$, 
  where $W_t$ is the wingspan, and $L_t$ the length for the specific aircraft
  type $t$
* aircraft area as viewed from the front, $a_F = W_t \times H_t = F_t$, 
  where $W_t$ is the wingspan, and $H_t$ the height for the specific aircraft
  type $t$

@@

The values $W_t$, $H_t$ and $L_t$ have been retrieved from [@performance_database;@skybrary;@doc8643].

@@

where 
* $E_{h,H}$ is the collision expectation in the hexagon $h$ at hour $H$,
* $W(h)$ is the median value of the weighting function over the area of hexagon
$h$,
* $\rho_{t,H}(h)$ is the traffic density for the \acr{ICAO} aircraft type $t$ in
cell $h$ at hour $H$, and
* $\varepsilon_t$ is effective exposed area for aircraft of type $t$.

@fig-collision depicts the determined hourly collision expectation.
As identified in @fig-density-summer2024-res-3 areas of higher traffic density observe a higher collision expectation.
<!-- WHAT ELSE CAN WE SAY ABOUT THIS PLOT? -->

@@

# Application Use Case: KOSMOS 482 Descent Craft re-entry
<!-- I would make this a separate section in the paper -->

In few days (at the time of writing of this paper) at 6h May 10, 2025
($\pm 2.8$ days), the KOSMOS 482 Descent Craft should reenter the
Earth's atmosphere [@kosmos482re-entry].
This is supposedly the lander module from a 1972 failed Soviet Venera mission [@SpaceReviewKosmos]
The module is reported to weigh about $\qty{480}{kg}$ and has a diameter of approximately $\qty{1}{m}$.
Being supposed to survive Venus's atmosphere which is much denser and hotter than Earth's atmosphere (surface pressure 93 times that of Earth and temperatures reaching \ang{467}C (\ang{872}F)),
it could re-enter Earth intact and not break into smaller fragments some of which could burn up.

@@

Once the reentry track is known, the H3 cells around \qty{35}{km} on each side 
can be identified to retrieve the values of collision expectancy.
It can then be decided whether it is advisable to close the airspace and
traffic around the area affected.

@@

<!-- opening statement linking back to abstract/intro and movtivation for this work --------->
Space objects and debris re-entry is an increasing risk to aviation and the general public.
The higher pace of commercial space operations will see an increase of launches and subsequent re-entries.
Dependent on the mission, objects re-entering the atmosphere can take on differnt forms, shapes, and sizes.
On top, a higher number of these entries may be uncontrolled.
With the continual growth of air traffic, there is a need to enhance aviation safety risk awareness and respond in a more tailored fashion to accomodate the associated uncertainties, reduce the impact on aviation, and limit the risk to the travelling public.

<!-- what was shown in the paper ------------------------------------------------------------>
This paper expands on previous work.
With a focus on Europe, we establish an upper bound for the traffic density based on actual flights.

... SOME MORE STUFF DONE IN THE PAPER ...

The findings are applied to a use-case: the imminent re-entry of the lander module of an earlier Venus mission.
The maximum collision expectation for the busiest day in 2024 is determined.

@@

The approach presented in this paper builds on previous research.
Our approach utilises H3-tesselation to reduce the spatial complexity.
This cell indexing approach reduces the computation significantly.

        WHAT ARE THINGS THAT ARE "WEAKNESSES" = STUFF THAT CAN BE IMPROVED? IT WOULD BE GOOD TO HAVE A LINE OR TWO ON THIS.

<!-- future work ---------------------------------------------------------------------------->

Future work will address the shortcomings. <!-- or can we say something about the next concrete steps to remedy some weaknesses -->

@@

As part of this research and to inform the decision-makers it is planned to explore the impact of megacostellations and flight growth forecast and what would happen in the mid- to long-term in varios scenarios of uncontrolled re-entries and the risk for European airspace.

>>>>

# Abstract
<!-- teaser --> The risk of collision between uncontrolled re-entering space debris and aircraft
is  increasing due to both ever-growing number of airborne flights and 
of orbiting bodies (and relevant re-entries). 
<!-- why important --> National and international authorities need to balance safety and economic
impact of adopted measures should a re-entry event be announced, usually on short
notice. <!-- no hard line breaks in abstracts -->
<!-- approach / experiments --> We employ hourly and daily spatial flight densities calculated from
trajectories as recorded by EUROCONTROL's Network Manager and
binned with the Uber H3 geospatial indexing system.  
Impact risk probabilities are then calculated based on previous theoretical
work. 
<!-- results / conclusions --> This work enhances the model by employing using actual flown flight trajectories as a basis for time-varying flight densities rather than relying on proxies such as the population density or ADS-B/Mode S message counts.
These densities combined with a buffered area along the eventual path of
re-entry, can provide mid-air impact probabilities to authorities as an aid
for deciding on safety measures such as airspace closures and/or re-routings.

@@

<!-- teaser opening --><!-- might need a rewrite - first jotting down -->
The quest for space is a driver for technological development.
Key milestones include the first human traval to space and surrounding the globe, the first and now many succeeding successful landings on the moon, and regular operations of space stations in the orbit.
These developments helped to identify options for the wider use of space, benefits, and LALALALA.
This is showcased by an increasing number of activities aimed to generate value through the exploration of space ranging from scientific understanding and experimentation, sustainable prooduction of goods in a non-gravity environment, services, and recently leisure.
The increased number of launches, however, also increased the number of artefacts <!-- falling down and --> re-entering the atmosphere.

<!-- zooming in on problem -->
With forecasts predicting a steadily growing number of flights, see @fig-forecast, and the modern
race to commercialise space, there is a need to re-evaluate the risks of
collision and casualties due to uncontrolled debris re-entries.
Overseeing authorities need methodologies and tools to assess which 
safety measures are the most adequate to minimise potential casualties and economic
impact.

Surging number of flights in Europe [@forecastUpdate2025] and in the
world, but more importantly under development/deployment mega-constellations
of satellites in \acr{LEO}, such as already active Starlink and OneWeb or
planned ones like Kuiper, Starshield, Xingwang et al. will add thousand
of orbiting objects with a short operational life. Some of these will 
inevitably re-enter the Earth's atmosphere in an uncontrolled manner, cfr. @tbl-megaconstellations [@megaconstellations].

@@

...

@@

We follow the same mathematical approach of 
[@patera2008; @wrightAirspaceClosuresDue2025] but instead of
approximating flight density with population density [@patera2008] or
aircraft transponder data [@wrightAirspaceClosuresDue2025], we use actual flown
flight trajectories.

The hourly flight density estimation is based on on flown \acr{IFR} trajectories as
recorded by EUROCONTROL's \acr{NM}[^1] and binned using Uber's H3 geospatial indexing system
[@uber-h3] in order to decrease spatial complexity which in turn enables fast processing.
This work will use H3 cells at resolution 3 which corresponds to hexagons of an
average area $\qty{12393}{km^2}$ and an average edge length of
$\qty{68.979}{km}$ [@h3resolutions] as depicted in @fig-cells.

@@

The flight densities are characterised on an hourly basis for all days in the busiest \acr{IATA} summer ISO week in 2023 and 2024.

This paper expands the works of [@wrightAirspaceClosuresDue2025; @patera2008] and
provides a reproducible example for States, Agencies or relevant authorities
to build upon.

@@

Next sections will develop the mathematical background for the evaluation of
collision and casualties risk of which traffic density is a key component.

@fig-density-summer2024-res-3 shows hourly flight density for the 5th of July 2024.
Expected outcomes such as cells at zero density for war or no-fly zones
(Ukraine, Libia or Syria) or typical traffic flows (North Atlantic, Canaries
and East-bound over Türkiye) are clearly highlighted.

@@

* constant debris fall speed $s_{d_k} = \qty{145}{mph} = S$ 
  [[pag 8 @patera2008]]
* constant aircraft speed, $s_{a} = C_t$, equal to its cruise speed, 
* aircraft area as viewed from the top, $a_T = W_t \times L_t = A_t$, 
  where $W_t$ is the wingspan, and $L_t$ the length for the specific aircraft
  type $t$
* aircraft area as viewed from the front, $a_F = W_t \times H_t = F_t$, 
  where $W_t$ is the wingspan, and $H_t$ the height for the specific aircraft
  type $t$

@@

The values $W_t$, $H_t$ and $L_t$ have been collected from [@performance_database;@skybrary;@doc8643]; we used \qty{1000}{m^2} for 
$\varepsilon_t$ in case of missing aircraft type.

@@

where $E_{h,H}$ is the collision expectation in the hexagon $h$ at hour $H$,
$W(h)$ is the median value of the weighting function over the area of hexagon
$h$,
$\rho_{t,H}(h)$ is the traffic density for the \acr{ICAO} aircraft type $t$ in
cell $h$ at hout $H$ and
$\varepsilon_t$ is effective exposed area for aircraft of type $t$.

@@

## Use Case: KOSMOS 482 Descent Craft re-entry

In few days (at the time of writing of this paper) at 6h May 10, 2025
($\pm 2.8$ days), the KOSMOS 482 Descent Craft should reenter the
Earth's atmosphere [@kosmos482re-entry].
This is supposedly the lander module from a 1972 failed Soviet Venera mission
[@SpaceReviewKosmos] of $\qty{480}{kg}$ weight and approximately
$\qty{1}{m}$ diameter.
Being supposed to survive Venus's atmosphere which is much denser and hotter than Earth's atmosphere (surface pressure 93 times that of Earth and temperatures reaching \ang{467}C (\ang{872}F)),
it could re-enter Earth intact and not break into smaller fragments some of which could burn up.

@@

Once the reentry track is known, the H3 cells around \qty{35}{km} on each side 
can be identified to retrieve the values of collision expectancy.
It can then be decided whether it is advisable to close the airspace and
traffic around the area afftected.

@@

<!-- opening statement linking back to abstract/intro and movtivation for this work --------->
Space debris re-entry is an increasing risk to aviation and the general public.

<!-- what was shown in the paper ------------------------------------------------------------>
This paper expands on previous work and ...

@@

@@

Future work will explore the impact of megacostellations and flight growth
forecast and what would happen in 20, 40, ..., 100 years in varios scenarios
of uncontrolled  reentries and the risk for European airspace.
